Reasonably Programmable Literal Notation

### 2018-06-21 icfp18/syntax-icfp18.tex
more edits to 2.1 and changed indexing notation to be consistent with relit

\noindent
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a discussion of contributions, limitations and future work.
% Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

%  We say more about LISP macros in Sec. \ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the supplement extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %Reason is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of Reason in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
% All examples written in Reason should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:Reason}.

@@

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  }
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer
      parser RegexParser.start
      in package regex_parser;
      dependencies = {module Regex = Regex}
    }
  }
\end{lstlisting}
\vspace{-5px}
\caption{The definition of \li{$regex}. Fig. \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}, which are detailed in Sec.~\ref{sec:provider-perspective}.}
\label{fig:regex-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{7px}
\begin{lstlisting}
  notation $regex = RegexNotation.$regex; /* or open RegexNotation */
  module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
  let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
  let restriction_template = (gene) => 
    $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
\end{lstlisting}
\vspace{-6px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tlm-example}
\end{subfigure}
\vspace{1px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
% \vspace{-4px}
\label{fig:regex-case-study}
\end{figure}

@@

The system enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on. In this case, Line 6 of Fig.~\ref{fig:regex-tlm-def} specifies that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally. In  general, the dependency can be an arbitrary module path, e.g. \li{module List = Core.Std.List}. The \li{Pervasives} module is implicitly opened in each expansion. All other bindings, both at the TLM application site and the TLM definition site, are not internally available to the expansion.

@@

From the client programmer's perspective, the benefit of this application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope, as they do when using unhygienic approaches (see Sec.~\ref{sec:existing-approaches}).%, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.

@@

\begin{figure}
\vspace{-10px}
\begin{displaymath}
\begin{array}{l}
  \text{\li{$regex `(GC$(DNA.any_base)GC)`}}\\[-10px]
  \xymatrix{
  ~ \ar[d]^{\text{The applied TLM lexes and parses the literal body to generate the \textbf{proto-expansion}.}} \\ ~
  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
  }\\[-5px]
  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}}\\[-10px]
  \xymatrix{
  ~ \ar[d]^{\text{The system performs proto-expansion validation and generates the \textbf{final expansion}.}} \\ ~
  % \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))}} \ar[d]^f\\ 
  }\\[-5px]
  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))}}
\end{array}
\end{displaymath}
\vspace{-4px}
\caption{Expansion of TLM applications occurs in two steps.}
\vspace{-6px}
\label{fig:expansion}
\end{figure}

@@

To address this problem, TLM parsers do not generate the final expansion directly, but rather a \emph{proto-expansion} that refers to spliced terms indirectly by location relative to the start of the provided literal body. For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as shown in the middle of Fig.~\ref{fig:expansion}. 
% \begin{lstlisting}[numbers=none]
%   Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))
% \end{lstlisting}
Here, \lismall{spliced<4..15 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because characters $4$ through $15$, zero-indexed and inclusive, of \lismall{GC$(DNA.any_base)GC} are \lismall{DNA.any_base}. We return to the type annotation on the splice reference, \li{Regex.t}, in Sec. \ref{sec:segment-typing} below. The context independence condition can be enforced straightforwardly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency in Fig.~\ref{fig:regex-tlm-def}, so all is well.

@@

\subsubsection{Segmentation}\label{sec:segmentation}
The finite set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body. The segmentation of the example above is the finite set containing one splice reference, \lismall{spliced<4..15 : Regex.t>}. For the more complex example from Line 5 of Fig.~\ref{fig:first-tlm-example}, the segmentation contains five splice references:
\begin{lstlisting}[numbers=none]
  { spliced<2..5 : Regex.t>, spliced<9..20 : Regex.t>, spliced<26..29 : string>, 
    spliced<33..44 : Regex.t>, spliced<49..52 : Regex.t> } 
\end{lstlisting}
The system checks that the segmentation does in fact segment the literal body, i.e. that the segments are in-bounds, of positive extent and non-overlapping. Adjacent spliced segments must also be separated by at least one character. The spliced segment locations can therefore be communicated unambiguously to the programmer by tools downstream of the expander, e.g. program editors and pretty printers, using secondary notation. In this paper, non-spliced segments are shown in color and spliced segments start in black.

When TLM applications are nested, a distinct color can be used at each depth. For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage. We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the supplement, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown. In Reason, \li{|>} is reverse function application.
\begin{figure}[t]
\begin{lstlisting}[numbers=none]
  $html `( SURL<div>
             <h3>Chemical Structure of Sucrose</h3>
             <$>EURL$smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` |> Smiles.to_svgSURL</$>
           </div>EURL )`
\end{lstlisting}
\caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
\label{fig:nesting-example}
\end{figure}

@@

Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool \cite{Merlin}, which is used by various Reason editor modes (e.g. for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor. Note that because the type is explicit, this information can be reported even when there is a parse or type error in a spliced expression. % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.

@@

\subsubsection{Capture}
In discussing the question of inferring a type for \li{gene} above, we neglected to consider one critical question: are we sure that the variable \li{gene} in the third spliced segment on Line 5 of Fig.~\ref{fig:first-tlm-example} is, in fact, a reference to the argument \li{gene} of the \li{restriction_template} function? After all, if we hold the expansion abstract then it may well be that the third spliced segment appears under (i.e. captures) a different binding of the identifier \li{gene}. For more common identifiers, e.g. \li{tmp}, inadvertent capture is not difficult to imagine. For example, consider this application site:
\begin{lstlisting}[numbers=none]
  let tmp = /* ... application site temporary ... */;
  $html `(SURL<h1><$>EURLf(tmp)SURL</$></h1>EURL)`;
\end{lstlisting}
Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
\begin{lstlisting}[numbers=none]
  let tmp = /* ... expansion-internal temporary ... */;
  Html.H1Element(tmp, spliced<7..12 : Html.t>);
\end{lstlisting}
If the final expansion was produced na\"ively, by syntactically replacing the splice reference with the final expansion recursively determined for the corresponding spliced expression, then the  variable \li{tmp} in the spliced expression would capture the expansion-internal binding of \li{tmp}. The result if the types of the two bindings differed would be a type error exposing the internal details of the expansion. If the types of the two bindings of \li{tmp} coincided, then there would be no static indication of the problem but there could be subtle and mysterious changes in run-time behavior. % Capture of invisible bindings is, due to these hazards, clearly unreasonable.

@@

%  Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}. Static functions  cannot refer to the surrounding variable bindings because those variables stand for run-time values. Instead, there is are separate \emph{static bindings} marked by the \li{static} keyword that populate a \emph{static environment} that is discarded after expansion finishes. A more detailed account of static evaluation, both informal and informal, is given in the supplement. An alternative design that allows for the explicit lowering of standard-phase modules to the static phase has also been proposed for OCaml \cite{Ocaml/macros}. %\ificfp We are closely following this work in our implementation. \else \fi
%   % return to in Sec. \ref{sec:static-eval}.

@@

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t;
  };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t;
  };
  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
  notation $re_type at Parsetree.core_type { /* ... (see text) ... */ };
  notation $proto_expr at ProtoExpr.t = $re_expr;
\end{lstlisting}
\caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-4px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

@@

The datatypes in OCaml's \li{Parsetree} module are necessarily intricate, given the sophistication of the OCaml system, so constructing expression encodings manually is generally too unwieldy. Addressing this class of problem is, of course, the motivation for this very paper, so it is only natural that we define TLMs  for working with OCaml parse trees using familiar surface syntax, extended with splice forms. In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}. The TLMs \li{$re_expr} and \li{$re_type} specified on Lines 10-11 of Fig.~\ref{fig:relit-util} implement quasiquotation for OCaml syntax trees using Reason's surface syntax. The supplement describes their implementation. For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future). We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}. Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}. Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms following the usual lexical scoping rules for \li{open}.

@@

\subsubsection{Other Implementation Strategies}
Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir: 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. For example, the supplement briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation. It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.

@@

% This explicit parameter passing discipline is reminiscent of work on explicit capture for distributed functions \cite{DBLP:conf/ecoop/MillerHO14}. By not implicitly giving expansions access to all definition-site bindings, we need not examine parse functions to reason about, e.g., renaming. Consequently, encodings of proto-terms can be values of standard datatypes (e.g. \li{proto_expr}) with variables represented simply as, e.g., strings, and quasiquotation notation can be expressing using TLMs (see supplement). A central design goal of Reason is to leave the OCaml semantics unchanged, so building quasiquotation in primitively, to integrate the free variables in term encodings into the overall binding discipline \cite{Bawd99a}, was in any case infeasible.

@@

\subsection{Core Language}\label{sec:s-XL}
Fig. \ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseBase$. This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The supplement shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g. $\unfold{e}$ rather than the $\aefoldp{p}$ pattern for values of recursive type.} Formally, core language terms are {abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}. This notational convention also clearly distinguishes core terms from unexpanded terms, which formally behave quite differently as we will describe below.

>>>>

\noindent
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a discussion of contributions, limitations and future work.
% Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the accompanying technical report. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

%  We say more about LISP macros in Sec. \ref{sec:existing-approaches}.% For the present purposes, we can consider Reason as essentially the calculus defined in the accompanying technical report extended with various conveniences that are commonly found in other ML-like languages and, notionally, orthogonal to TLMs. %Reason is, as its name suggests, a conceptual descendent of ML. It diverges from other dialects of ML that have a similar type structure in that it has a bidirectional type system \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala \cite{OdeZenZen01}) for reasons that have to do with the mechanism of TLM implicits described in Chapters \ref{chap:tsls} and \ref{chap:ptsms}. 
%The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal definition of Reason in this work is both to emphasize that the primitives we introduce are ``insensitive'' to the details of the underlying type structure of the language (so TLMs can be considered for inclusion in a variety of languages, not only dialects of ML), and to avoid distracting the reader (and the author) with definitions that are already well-understood in the literature and that are orthogonal to those that are the focus of this work. 
% We will not formally define these features mainly to avoid unnecessarily complicating our presentation with details that are not essential to the ideas presented herein. As such, 
% All examples written in Reason should be understood to be informal motivating material for the subsequent formal material. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:Reason}.

@@

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = Empty
           | AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  };
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-4px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation = {
    notation $regex at Regex.t {
      lexer  RegexLexer
      parser RegexParser.start
      in package regex_parser;
      dependencies = {module Regex = Regex}
    };
  }
\end{lstlisting}
\vspace{-5px}
\caption{The definition of \li{$regex}. Fig. \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}, which are detailed in Sec.~\ref{sec:provider-perspective}.}
\label{fig:regex-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{7px}
\begin{lstlisting}
  notation $regex = RegexNotation.$regex; /* or open RegexNotation */
  module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
  let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
  let restriction_template = (gene) => 
    $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
\end{lstlisting}
\vspace{-4px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tlm-example}
\end{subfigure}
\vspace{1px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
% \vspace{-4px}
\label{fig:regex-case-study}
\end{figure}

@@

The system enforces a strong context independence condition on generated expansions by requiring that the TLM definition explicitly specify all modules that the expansions it generates might internally depend on. In this case, Line 6 of Fig.~\ref{fig:regex-tlm-def} specifies that generated expansions might use the module \li{Regex}, again as it is bound at the TLM definition site, using the module variable \li{Regex} internally. In  general, the dependency can be an arbitrary module path, e.g. \li{module List = Core.Std.List}. The \li{Pervasives} module is implicitly opened in each expansion. All other bindings, whether at the TLM application site or the TLM definition site, are not internally available to the expansion.

@@

From the client's perspective, the benefit of this application site context independence discipline is clear---clients do not need to give any thought to which bindings the expansion might invisibly be assuming are in scope, as they do when using unhygienic approaches (see Sec.~\ref{sec:existing-approaches}).%, e.g., \li{camlp4} \cite{de2003camlp4}, Sugar* \cite{erdweg2011sugarj}, PPX rewriters \cite{ocaml-manual} or Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s}.

@@

\begin{figure}
\vspace{-10px}
\begin{displaymath}
\begin{array}{l}
  \text{\li{$regex `(GC$(DNA.any_base)GC)`}}\\[-10px]
  \xymatrix{
  ~ \ar[d]^{\text{The applied TLM lexes and parses the literal body to generate the \textbf{proto-expansion}.}} \\ ~
  }\\[-5px]
  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..16 : Regex.t>, Regex.Str "GC"))}}\\[-10px]
  \xymatrix{
  ~ \ar[d]^{\text{The system performs proto-expansion validation and recursively expands the splices to generate the final \textbf{expansion}.}} \\ ~
  }\\[-5px]
  \text{\li{Regex.Seq(Regex.Str "GC", Regex.Seq(DNA.any_base, Regex.Str "GC"))}}
\end{array}
\end{displaymath}
\vspace{-4px}
\caption{TLM applications expand in two steps.}
\vspace{-6px}
\label{fig:expansion}
\end{figure}

@@

To address this problem, TLM parsers do not generate the final expansion directly, but rather a \emph{proto-expansion} that refers to spliced terms indirectly by location relative to the start of the provided literal body. For example, the proto-expansion generated by \li{$regex} for the example above can be pretty printed as shown in the middle of Fig.~\ref{fig:expansion}. 
% \begin{lstlisting}[numbers=none]
%   Regex.Seq(Regex.Str "GC", Regex.Seq(spliced<4..15 : Regex.t>, Regex.Str "GC"))
% \end{lstlisting}
Here, \lismall{spliced<4..16 : Regex.t>} is a reference to the spliced expression \li{DNA.any_base} because the zero-indexed subsequence $[4, 16)$ of \lismall{GC$(DNA.any_base)GC} is \lismall{DNA.any_base}. We return to the type annotation on the splice reference, \li{Regex.t}, when we discuss \textbf{Segment Typing} in Sec. \ref{sec:segment-typing} below. The context independence condition can be enforced directly on the proto-expansion---the only free variable in the proto-expansion is \li{Regex}, which is an explicitly listed dependency in Fig.~\ref{fig:regex-tlm-def}, so all is well.

@@

\subsubsection{Segmentation}\label{sec:segmentation}
The finite set of splice references in the proto-expansion generated for a literal body is called the \emph{segmentation} of that literal body. The segmentation of the example above is the finite set containing one splice reference, \lismall{spliced<4..16 : Regex.t>}. For the more complex example from Line 5 of Fig.~\ref{fig:first-tlm-example}, the segmentation contains five splice references:
\begin{lstlisting}[numbers=none]
  { spliced<2..6 : Regex.t>, spliced<9..21 : Regex.t>, spliced<26..30 : string>, 
    spliced<33..45 : Regex.t>, spliced<49..53 : Regex.t> } 
\end{lstlisting}
The system checks that the segmentation does in fact segment the literal body, i.e. that the segments are in-bounds, of positive extent and non-overlapping. Adjacent spliced segments must also be separated by at least one character. The spliced segment locations can therefore be communicated unambiguously to the programmer by tools downstream of the expander, e.g. program editors and pretty printers, using secondary notation. In this paper, non-spliced segments are shown in color and spliced segments start in black.

When TLM applications are nested, a distinct color can be used at each depth. For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage. We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the accompanying technical report, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown. In Reason, \li{|>} is reverse function application.
\begin{figure}[t]
\begin{lstlisting}[numbers=none]
  $html `( SURL<div>
             <h3>Chemical Structure of Sucrose</h3>
             <$>EURL$smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` |> Smiles.to_svgSURL</$>
           </div>EURL )`
\end{lstlisting}
\vspace{-2px}
\caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
\label{fig:nesting-example}
\end{figure}

@@

Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool\todo{cite ICFP merlin paper}{} \cite{Merlin}, which is used by various Reason editor modes (e.g. for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor. Note that because the type is explicit, this information can be reported even when there is a parse or type error in a spliced expression. % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.

@@

\subsubsection{Capture}
In discussing the question of inferring a type for \li{gene} above, we neglected to consider one critical question: are we sure that the variable \li{gene} in the third spliced segment on Line 5 of Fig.~\ref{fig:first-tlm-example} is, in fact, a reference to the argument \li{gene} of the \li{restriction_template} function? After all, if we hold the expansion abstract then it may well be that the third spliced segment appears under (i.e. captures) a different binding of the identifier \li{gene}. For more common identifiers, e.g. \li{tmp}, inadvertent capture is not difficult to imagine. For example, consider this application site:
\begin{lstlisting}[numbers=none]
  let tmp = /* ... application site temporary ... */;
  $html `(SURL<h1><$>EURLf(tmp)SURL</$></h1>EURL)`;
\end{lstlisting}
Now consider the scenario where the proto-expansion generated by \li{$html} has the following form:
\begin{lstlisting}[numbers=none]
  let tmp = /* ... expansion-internal temporary ... */;
  Html.H1Element(tmp, spliced<7..13 : Html.t>);
\end{lstlisting}
If the final expansion was produced na\"ively, by syntactically replacing the splice reference with the final expansion recursively determined for the corresponding spliced expression, then the  variable \li{tmp} in the spliced expression would capture the expansion-internal binding of \li{tmp}. The result if the types of the two bindings differed would be a type error exposing the internal details of the expansion. If the types of the two bindings of \li{tmp} coincided, then there would be no static indication of the problem but there could be subtle and mysterious changes in run-time behavior. % Capture of invisible bindings is, due to these hazards, clearly unreasonable.

@@

%  Because the parse function is applied during this phase, rather than at run-time, we call it a \emph{static function}. Static functions  cannot refer to the surrounding variable bindings because those variables stand for run-time values. Instead, there is are separate \emph{static bindings} marked by the \li{static} keyword that populate a \emph{static environment} that is discarded after expansion finishes. A more detailed account of static evaluation, both informal and informal, is given in the accompanying technical report. An alternative design that allows for the explicit lowering of standard-phase modules to the static phase has also been proposed for OCaml \cite{Ocaml/macros}. %\ificfp We are closely following this work in our implementation. \else \fi
%   % return to in Sec. \ref{sec:static-eval}.

@@

\begin{figure}
\begin{lstlisting}[deletekeywords={spliced}]
  module Segment : { 
    type t = {start_pos: int, end_pos: int};
    let read_to : string => Lexing.lexbuf => Segment.t;
  };
  exception ExpansionError({msg: string, loc: option(Segment.t)});
  module ProtoExpr : { 
    type t = Parsetree.expression; /* from the OCaml compiler library */
    let spliced : Segment.t => Parsetree.core_type => t;
  };
  notation $re_expr at Parsetree.expression { /* ... (see text) ... */ };
  notation $proto_expr at ProtoExpr.t = $re_expr;
\end{lstlisting}
\caption{A fragment of the \li{Relit} module signature relevant to expression TLM providers.}
\label{fig:relit-util}
\vspace{-4px}
\end{figure}
    % let match_spliced : t => option(Segment.t, ProtoTyp.t);

@@

The datatypes in OCaml's \li{Parsetree} module are necessarily intricate, given the sophistication of the OCaml system, so constructing expression encodings manually is generally too unwieldy. Addressing this class of problem is, of course, the motivation for this very paper, so it is only natural that we define TLMs  for working with OCaml parse trees using familiar surface syntax, extended with splice forms. In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}. The TLM \li{$re_expr}  specified on Line 10 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax. The accompanying technical report describes its implementation. For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future). We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}. Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}. Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms following the usual lexical scoping rules for \li{open}.

@@

\subsubsection{Other Implementation Strategies}
Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir: 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. For example, the accompanying technical report briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation. It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.

@@

% This explicit parameter passing discipline is reminiscent of work on explicit capture for distributed functions \cite{DBLP:conf/ecoop/MillerHO14}. By not implicitly giving expansions access to all definition-site bindings, we need not examine parse functions to reason about, e.g., renaming. Consequently, encodings of proto-terms can be values of standard datatypes (e.g. \li{proto_expr}) with variables represented simply as, e.g., strings, and quasiquotation notation can be expressing using TLMs (see accompanying technical report). A central design goal of Reason is to leave the OCaml semantics unchanged, so building quasiquotation in primitively, to integrate the free variables in term encodings into the overall binding discipline \cite{Bawd99a}, was in any case infeasible.

@@

\subsection{Core Language}\label{sec:s-XL}
Fig. \ref{fig:U-expanded-terms} gives the syntax of the core language, $\miniVerseBase$. This language forms a standard pure functional language with partial function types, quantification over types, recursive types, labeled product types, labeled sum types and support for pattern matching.\footnote{The accompanying technical report shows how to remove pattern matching  (and pattern TLMs) from the calculus by adding the usual elimination forms, e.g. $\unfold{e}$ rather than the $\aefoldp{p}$ pattern for values of recursive type.} Formally, core language terms are {abstract binding trees} (ABTs) identified up to alpha-equivalence, so we follow the syntactic conventions of \citet{pfple1}. This notational convention also clearly distinguishes core terms from unexpanded terms, which formally behave quite differently as we will describe below.
